Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2018-06-15 paper.tex
Remove last four TODO notes
* 'look at Bretts tests and make clearer statement'
* 'give eqn for L and avg eps'
* 'improve; this is too simplistic'
* 'also explain why G_nue > G_nuebar for theta=0 at cosA>0.995'

Ignore these for the time being, and come back to them if the referee
is concerned.

As Fig.~\ref{fig:mfps-50km} reveals, integrating the distribution functions
over the energies $\varepsilon\in(0,100)$~MeV
probes the numerical solution over length scales from
0.1~km to $10^{5}\,{\rm km}$.
Fig.~\ref{fig:homogeneous_isotropic} shows the error in our results for
the \emph{noscat} case.
In this simple case, the dominant source of error is the neglected
boundary term with a relative scale of $e^{-\tau_{\rm term}}\approx10^{-6}$.
In the following inhomogeneous configurations,
errors from the integration dominate over the boundary term.
In this test and only this test, we used a higher-order integrator, a 5th
order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed}. This is because
\footnote{TODO (Matt, Brett): look at Brett's tests and make clearer statement}
our default 3rd order Runge-Kutta algorithm estimates a vanishing error
in this configuration.

@@

The total luminosities and average energies
measured by the different treatments are
presented in Tab.~\ref{tab:collapse_rt_vs_mc}.
\footnote{TODO (Brett): give equation for L and avg eps.}
As expected, $\nu_e$ is least affected by scattering, and $\nu_x$ most.
In fact, without scattering, $\nu_x$ luminosities are overestimated by more than
two orders of magnitude, due to the steep increase in temperature with depth
in the inner core.
Though contributing only a fraction of the total luminosity,
the average energy of neutrinos scattered to the observer from the envelope
outside the shock, $\langle\varepsilon\rangle^{\rm sc}$,
is poorly characterized by the \emph{noscat} treatment for all species.
By contrast, both the \emph{gray} and \emph{spectral} \emph{scat} treatments
faithfully describe the high scattered energies from the envelope.

@@

Though the ray tracing sampling of $f$ is done in full general relativity,
in our calculation of moments at the observer positions,
we make the simplifying assumption of Minkowski spacetime.
This assumption introduces errors in our moments of order
$GM/r=4\,{\rm km}/250\,{\rm km}$
\footnote{TODO (Brett, Francois, Matt): improve; this is too simplistic}
(with $G$ the gravitational constant and $M$ the central mass),
or $\sim 1\%$.

@@

In Figs.~\ref{fig:nsns_cosA_distrib_G_00deg}
and \ref{fig:nsns_cosA_distrib_G_90deg}, we show the distributions of neutrino
number density over incoming angle
for the observer on the polar axis $\theta=0^\circ$,
and in the equatorial plane $\theta=90^\circ$,
defined analagously to Eqn.~\ref{eqn:J_per_cosA}.
As in the case of the collapse profile in Sec.~\ref{ssec:test_collapse},
the dominant effect of elastic scattering on $\tilde{G}(\cos A)$
is to spread the distribution out to larger angles, by generally decreasing
the number of neutrinos coming from the core while increasing the number
of neutrinos coming from the disk.
\footnote{TODO (Evan, Brett): also explain why $G_{\nu_e}>G_{\bar{\nu}_e}$
  for $\theta=0$ at $\cos A>0.995$.}

@@

\begin{figure}
  %\includegraphics[width=\columnwidth]{cosA_distrib-intensity-250km-00deg}
  \resizebox{\columnwidth}{!}{\input{cosA_distrib-G-250km-00deg.tex}}
  % Data:   $DIR/Lev2-e30a150b20-{,no}scat-04
  % Script: $DIR/plot-angular_distribution-number_density-compare.gnu
  % w/ DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20170324-ray_tracing_nsns/Run03/000Mo-000.0deg-250km
  \caption{Distribution of number density $\tilde{G}(\cos A)$ over incoming angle,
    defined analagously to Eqn.~\ref{eqn:J_per_cosA}.
    The integral of this quantity over $d\cos A$ gives the total number density
    measured by this observer on the rotation axis.
    Volume data from the hypermassive neutron star configuration
    (presented in Sec.~\ref{ssec:test_disk_comparison}).}
  \label{fig:nsns_cosA_distrib_G_00deg}
\end{figure}

>>>>

As Fig.~\ref{fig:mfps-50km} reveals, integrating the distribution functions
over the energies $\varepsilon\in(0,100)$~MeV
probes the numerical solution over length scales from
0.1~km to $10^{5}\,{\rm km}$.
Fig.~\ref{fig:homogeneous_isotropic} shows the error in our results for
the \emph{noscat} case.
In this simple case, the dominant source of error is the neglected
boundary term with a relative scale of $e^{-\tau_{\rm term}}\approx10^{-6}$.
In the following inhomogeneous configurations,
errors from the integration dominate over the boundary term.
In this test and only this test, we used a higher-order integrator, a 5th
order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed}. This is because
our default 3rd order Runge-Kutta algorithm estimates a vanishing error
in this configuration.

@@

The total luminosities and average energies
measured by the different treatments are
presented in Tab.~\ref{tab:collapse_rt_vs_mc}.
As expected, $\nu_e$ is least affected by scattering, and $\nu_x$ most.
In fact, without scattering, $\nu_x$ luminosities are overestimated by more than
two orders of magnitude, due to the steep increase in temperature with depth
in the inner core.
Though contributing only a fraction of the total luminosity,
the average energy of neutrinos scattered to the observer from the envelope
outside the shock, $\langle\varepsilon\rangle^{\rm sc}$,
is poorly characterized by the \emph{noscat} treatment for all species.
By contrast, both the \emph{gray} and \emph{spectral} \emph{scat} treatments
faithfully describe the high scattered energies from the envelope.

@@

Though the ray tracing sampling of $f$ is done in full general relativity,
in our calculation of moments at the observer positions,
we make the simplifying assumption of Minkowski spacetime.
This assumption introduces errors in our moments of order
$GM/r=4\,{\rm km}/250\,{\rm km}$
(with $G$ the gravitational constant and $M$ the central mass),
or $\sim 1\%$.

@@

In Figs.~\ref{fig:nsns_cosA_distrib_G_00deg}
and \ref{fig:nsns_cosA_distrib_G_90deg}, we show the distributions of neutrino
number density over incoming angle
for the observer on the polar axis $\theta=0^\circ$,
and in the equatorial plane $\theta=90^\circ$,
defined analogously to Eqn.~\ref{eqn:J_per_cosA}.
As in the case of the collapse profile in Sec.~\ref{ssec:test_collapse},
the dominant effect of elastic scattering on $\tilde{G}(\cos A)$
is to spread the distribution out to larger angles, by generally decreasing
the number of neutrinos coming from the core while increasing the number
of neutrinos coming from the disk.

@@

\begin{figure}
  %\includegraphics[width=\columnwidth]{cosA_distrib-intensity-250km-00deg}
  \resizebox{\columnwidth}{!}{\input{cosA_distrib-G-250km-00deg.tex}}
  % Data:   $DIR/Lev2-e30a150b20-{,no}scat-04
  % Script: $DIR/plot-angular_distribution-number_density-compare.gnu
  % w/ DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20170324-ray_tracing_nsns/Run03/000Mo-000.0deg-250km
  \caption{Distribution of number density $\tilde{G}(\cos A)$ over incoming angle,
    defined analogously to Eqn.~\ref{eqn:J_per_cosA}.
    The integral of this quantity over $d\cos A$ gives the total number density
    measured by this observer on the rotation axis.
    Volume data from the hypermassive neutron star configuration
    (presented in Sec.~\ref{ssec:test_disk_comparison}).}
  \label{fig:nsns_cosA_distrib_G_00deg}
\end{figure}
